---
title: "Backend and provenance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backend and provenance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`cudaembedr` keeps upstream work separate from the work performed by the
current embedding:

- **source provenance** describes the input object, such as the stages used to
  compute a PCA result;
- **current compute provenance** describes the embedding stages returned by
  `cuda_provenance()`.

This distinction prevents a CUDA source from being mistaken for an
end-to-end CUDA embedding.

## A runnable CPU diffusion map

This example computes PCA and the diffusion map on the CPU. Row names are
preserved through both results.

```{r cpu-diffusion}
library(cudaembedr)

set.seed(1)
observations <- matrix(rnorm(40 * 5), 40, 5)
rownames(observations) <- paste0("observation_", seq_len(nrow(observations)))

pca <- cudalearnr::cuda_pca(
  observations,
  n_components = 3,
  device = "cpu"
)
embedding <- cuda_diffusion_map(
  pca,
  n_components = 2,
  device = "cpu"
)

embedding
head(embedding_coordinates(embedding))
embedding$source_device
embedding$source_compute_device
embedding$source_provenance
cuda_provenance(embedding)
```

`embedding$source_provenance` contains the upstream PCA stages.
`cuda_provenance(embedding)` contains only the current distance, kernel, and
eigendecomposition stages. With `device = "cpu"`, all three current stages use
the CPU and `embedding$compute_device` is `"cpu"`.

UMAP and t-SNE use fixed CPU backends (`uwot` and `Rtsne`, respectively).
Their source metadata can still describe an upstream CUDA result, but their
current embedding stage remains CPU.

## Why a CUDA diffusion map is hybrid

`device` in `cuda_diffusion_map()` controls the pairwise-distance stage.
Kernel construction and eigendecomposition are currently fixed to the CPU.
The truthful stage summary is therefore:

| Distance stage | Kernel stage | Eigendecomposition stage | `compute_device` |
|---|---|---|---|
| CPU | CPU | CPU | `"cpu"` |
| CUDA | CPU | CPU | `"hybrid"` |

The result's top-level `backend` names the eigendecomposition backend
(`base-eigen` or `RSpectra`). Use `cuda_provenance()` when the backend and
device of every stage matter.

## Quadratic memory is still the limiting factor

Diffusion maps currently materialize a complete pairwise-distance matrix,
followed by dense kernel and normalized matrices. One double-precision
`n`-by-`n` matrix alone requires approximately `8 * n^2` bytes:

```{r memory-estimate}
one_dense_matrix_gib <- function(n) {
  8 * as.double(n)^2 / 1024^3
}

data.frame(
  observations = c(1e3, 1e4, 5e4),
  one_dense_matrix_GiB = round(
    one_dense_matrix_gib(c(1e3, 1e4, 5e4)),
    2
  )
)
```

Peak memory is several times this single-matrix estimate because multiple
dense intermediates coexist. CUDA can accelerate distance computation, but it
does not remove the `O(n^2)` host-memory requirement of the current
implementation. Estimate memory before running large data and prefer methods
that do not materialize all pairwise distances when `n` is large.

## Optional CUDA execution and the hardware gate

The next block runs only when a usable CUDA backend exists. It requests CUDA
explicitly and verifies the hybrid contract: CUDA distance, CPU kernel, and CPU
eigendecomposition.

The cudaverse NVIDIA workflow sets `CUDAVERSE_REQUIRE_CUDA=true`. Under that
gate, an unavailable CUDA runtime is an error rather than an allowed skip.

```{r optional-cuda}
require_cuda <- identical(
  tolower(Sys.getenv("CUDAVERSE_REQUIRE_CUDA", "false")),
  "true"
)
cuda_ready <- cudatensr::cuda_available()

if (require_cuda && !cuda_ready) {
  stop("The hardware gate requires a usable CUDA backend.")
}

if (cuda_ready) {
  embedding_cuda <- cuda_diffusion_map(
    observations,
    n_components = 2,
    device = "cuda"
  )
  provenance_cuda <- cuda_provenance(embedding_cuda)
  stage_devices <- stats::setNames(
    provenance_cuda$device,
    provenance_cuda$stage
  )

  stopifnot(
    identical(embedding_cuda$compute_device, "hybrid"),
    identical(unname(stage_devices["distance"]), "cuda"),
    identical(unname(stage_devices["kernel"]), "cpu"),
    identical(unname(stage_devices["eigendecomposition"]), "cpu")
  )

  provenance_cuda
}
```

Use `device = "cuda"` when CUDA is mandatory. Use `device = "auto"` only when
the recorded CPU fallback is acceptable. In either case,
`cuda_provenance()` is the authoritative record of what actually ran.
